\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-11-02 paper.tex
tweak

\section{Event ARGs}
\label{sec-eARG}
% Trying to be totally, boringly, abundantly, blatantly clear here what we're
% talking about.
In this section we define the classical view of an ARG data structure,
and illustrate its limitations. We are
interested in the details of how ARGs are described mathematically,
and as a consequence, how they are represented in a practical sense
as the output of inference programs.
Where details of an ARG data structure (the encoding) are
% Three papers 10 years apart will do here, right?
provided~\citep[e.g.][]{
wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability}
they follow the approach described by Griffiths and colleagues
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions), and a large number of
% Note the "an" here to avoid cases like ARGweaver
ARG inference methods use it as an output format
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
% Griffiths never mentions "sampling events", so let's not get
% into that. It's not an important detail, and we don't want to be
% overly precise about something that people haven't actually written
% down.
In this Griffiths encoding we have two types of internal node in the graph,
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints. These breakpoints
are used to construct the local trees.
This is done by tracing pastwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Because it is focused on recording events and their properties,
we will refer to this Griffiths encoding as the ``event ARG'' or eARG
encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
We assume here that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

In eukaryotes, ARGs are a result of the cellular processes of mitosis
(which leads to common ancestry and coalescence)
and meiosis (which leads to recombination, including both crossovers and gene conversion).
Fig.~\ref{fig-simplification} shows a schematic of the events (coloured red and blue squares) and
the genomes (chromosome icons) that occur in the cellular germline of a simplified, diploid multicellular
hermaphrodite eukaryote with partially overlapping generations.
% It may be worth putting in something here about events being a slippery concept
% (we have already mentioned this at the end of Section 3
Here, an event is not represented by a specific
genome. Rather, genomes can be associated with (or ``tag'') events above or below them (and as argued in
Section~\ref{sec-precision}, the gARG representation allows these associated genomes to be arbitrarily distant
in time or number of cell
divisions from the event itself). For example, tagging the two genomes above a recombination event leads to the
2 node representation seen in Figs.~\ref{fig-simplification} and \ref{hudson_vs_bigARG}, whereas tagging the genome
below a recombination event leads to the more conventional graphs in Figs.~\ref{fig-ancestry-resolution} and \ref{fig-inferred-args}A,B.

Thick black lines trace an ARG that describes the ancestry of the lower arm of the
four sampled chromosomes, and these lines pass through gARG nodes (green chromosomal regions).
A number of these nodes have only one parent and one child in the ARG; these would often be omitted
(taken as implicit) in a minimally simplified gARG.

In particular, this schematic illustrates a case where three ARG lineages coalesce within
a single individual ($D_{10}$), corresponding to what is normally represented by a polytomy
(e.g. Fig.~\ref{fig-simplification} node \noderef{k}). The process that generates
these three lineages actually consists of two successive bifurcations, reflecting the fact that
the only known method of reproducing DNA is by (semi-conservative) duplication. Under this schematic,
polytomies therefore represent a gARG-enabled merging of multiple coalescent events into an older
node, just as ARG nodes with more than 2 parents represent a merging of multiple recombination
events into a younger node.

Note that mutations can occur along any lineage in Fig.~\ref{fig-cell-lines}. This means, for example,
that a mutation in the first cell division of $D_{10}$ could be shared between
the two gametes produced by the cells in the left half of $D_{10}$ but not shared by
the right hand gamete. With a high enough mutation rate (or equivalently, a long enough genome), each
round of mitotic germline genome duplication within a single multicellular organism could
potentially be distinguished.
This implies that in ARGs that have been inferred from real data, polytomies represent a lack
of mutational information (rather than marking a single ``multifurcation event'').
The ability to represent polytomies, and conceptualise them as a (deliberately) imprecise statement about the order
of coalescence within and between individuals,
fits naturally into the gARG framework presented in this paper.

>>>>

\section{Event ARGs}
\label{sec-eARG}
% Trying to be totally, boringly, abundantly, blatantly clear here what we're
% talking about.
In this section we define the classical view of an ARG data structure,
and illustrate its limitations. We are
interested in the details of how ARGs are described mathematically,
and as a consequence, how they are represented in a practical sense
as the output of inference programs.
Where details of an ARG data structure (the encoding) are
% Three papers 10 years apart will do here, right?
provided~\citep[e.g.][]{
wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability}
they follow the approach described by Griffiths and colleagues
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions), and a large number of
% Note the "an" here to avoid cases like ARGweaver
ARG inference methods use it as an output format
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
% Griffiths never mentions "sampling events", so let's not get
% into that. It's not an important detail, and we don't want to be
% overly precise about something that people haven't actually written
% down.
In this Griffiths encoding we have two types of internal node in the graph,
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints. These breakpoints
are used to construct the local trees.
This is done by tracing pastwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Because it is focused on recording events and their properties,
we will refer to this Griffiths encoding as the ``event ARG'' or eARG
encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
We assume here that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
% Some eARG implementations incorporate additional information to
% distinguish left and right parents into the
% encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.

@@

In eukaryotes, ARGs are a result of the cellular processes of mitosis and meiosis.
Mitosis leads to common ancestor events,
and meiosis leads to recombination events (both crossover and gene conversion).
Fig.~\ref{fig-simplification} shows a schematic of the events  and
the genomes (chromosome icons) that occur in the cellular
germline of a simplified, diploid multicellular
hermaphrodite eukaryote with partially overlapping generations.
Here, an event is not represented by a specific
genome. Rather, genomes can be associated with, or ``tag'', events above
(ancestral to) or below (descended from) them.
For example, tagging the two genomes above a recombination event leads to the
two node representation seen in
Figs.~\ref{fig-simplification} and \ref{hudson_vs_bigARG},
whereas tagging the genome
below a recombination event leads to the more
conventional graphs in Figs.~\ref{fig-ancestry-resolution} and \ref{fig-inferred-args}A,B.

%% JK:  What does this add that is not in caption?
% Thick black lines trace an ARG that describes the ancestry of the lower arm of
% the four sampled chromosomes, and these lines pass through gARG nodes (green
% chromosomal regions). A number of these nodes have only one parent and one
% child in the ARG; these would often be omitted (taken as implicit) in a
% minimally simplified gARG.

The schematic illustrates an important point about the biological reality of polytomies.
Three linages coalesce in the left-hand genome of individual $D_{10}$, but do
so as the result of two successive bifurcations. This is \emph{necessarily} so,
because the only known method of reproducing DNA is by (semi-conservative) duplication.
Whether this polytomy is resolvable depends on the available mutational data.
Mutations can occur along any cell lineages. For example,
a mutation in the first cell division of $D_{10}$ could be shared between
the two gametes produced by the cells in the left half of $D_{10}$
but not shared by the right hand gamete.
With enough mutations, each
round of mitotic germline genome duplication within a single multicellular organism could
in principle be distinguished.

% Note that mutations can occur along any lineage in Fig.~\ref{fig-cell-lines}. This means, for example,
% that a mutation in the first cell division of $D_{10}$ could be shared between
% the two gametes produced by the cells in the left half of $D_{10}$ but not shared by
% the right hand gamete. With a high enough mutation rate (or equivalently, a long enough genome), each
% round of mitotic germline genome duplication within a single multicellular organism could
% potentially be distinguished.
% This implies that in ARGs that have been inferred from real data, polytomies represent a lack
% of mutational information (rather than marking a single ``multifurcation event'').
% The ability to represent polytomies, and conceptualise them as a (deliberately) imprecise statement about the order
% of coalescence within and between individuals,
% fits naturally into the gARG framework presented in this paper.
